Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-07-24 paper.tex
Started explaning equation for multiple reads. Stopped because equation seems to be incorrect.
Updates in description of cluster reads

\title{Sequential algorithms to split and merge terabyte-scale 3D images}

@@

Three-dimensional images that exceed typical memory size are
increasingly found in a variety of disciplines. BigBrain, for
instance, is a 3D histological image of the human brain that
represents 1~TB of data organized in 3600 planes at full resolution
(1$\mu$m in-plane), and 76~GB at a 40$\mu$m isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. Other
examples found in medical imaging, our primary domain of interest,
include \todo{electromicroscopy,spectroscopy-francoise-peyrin}. As
such images would typically be processed on a computing cluster,
possibly using locality-aware file systems such as HDFS, software
libraries are needed to split and merged such images efficiently. We
introduce and compare a family of algorithms to address this problem
in a sequential environment made of a single storage disk. We are
aware of the need for parallel algorithms that would split and merge
images to and from an array of disks but we plan to describe such
algorithms in a subsequent paper based on the results of the present
one.

@@

We adopt the following notations:
\begin{itemize}
\item $R=D^3$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel in the reconstructed image and chunks (in B).
\item $n$: number of chunks (blocks or slices).
\item $\alpha_r$: data read rate from disk (in B/s).
\item $\alpha_w$: data write rate to disk (in B/s).
\item $\beta$: disk access time (in s).
\item $\sigma: \mathbb{R}^{+} \rightarrow \mathbb{R}^{+}$: seek time (in s), function of the seek
  distance (in B).
\item $m$: amount of available RAM (in B).
\end{itemize}

@@

As detailed in Algorithm~\ref{algo:cluster-reads}, cluster reads load
multiple blocks in memory, concatenate them in a buffer and write the
buffer sequentially in the reconstructed image. The buffer is a
slightly complex data structure, for instance an associative array or
a Python dictionary, capable of storing multiple disjoint sequences of
contiguous bytes without having to allocate memory for the bytes
between such sequences. Figure~\ref{fig:cluster-reads-buffer}
illustrates how the buffer would fill up when two blocks of one slice
and three rows per slice are loaded. Implementation details about the
buffer are given later.
\begin{figure}
\centering
\def\svgwidth{\columnwidth}
\input{figures/buffer-cluster-reads.pdf_tex}
\caption{Illustration of the buffer used in cluster reads. White
  portions are not allocated. [a,b] refers to block a, row b.}
\label{fig:cluster-reads-buffer}
\end{figure}
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}
    \STATE list = sort blocks by increasing (x,y,z) values
    \STATE initialize buffer
    \STATE index = 0
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m}
        \STATE block = list[index]
        \STATE read block and insert it in buffer
        \STATE index += 1
        \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

The number of seeks performed by cluster reads depends on how blocks
loaded in memory arrange in the reconstructed image. In the best case,
complete contiguous slices of the reconstructed image can be assembled
in memory and written without any seeking. In the worst case, some
rows in the reconstructed image are not covered by blocks loaded in
memory: O($d^2$) seeks are then required to write the memory load, one
for every partial row in every partial slice. In the intermediary
case, rows are complete but some slices can only be partially
reconstructed: O($d$) seeks are then
required. We have:
\begin{equation}
  N_\mathrm{cluster\_reads}= n + \sum_{i=1}^{\ceil*{\frac{bR}{m}}}f_i
\end{equation}
where $f_i$ is the number of seeks done in the $i^{th}$ memory
load. Figure~\ref{fig:cluster-reads} lists the possible block
arrangements and the corresponding values of $f_i$. \todo{Val\'erie: Check that
  ordering is consistent with Nifti's.}
%% In case (1) on
%% this Figure, the memory load completely fills rows and slices in the
%% reconstructed image. Thus it can be written with a single seek,
%% $f_i$=1. In case (2), all the rows are complete but $d$ slices are
%% incomplete; one seek is required for each slices, $f_i=d$. In case
%% (3), $d$ rows of $d$ slices are incomplete, $f_i=d^2$. In case (4),
%% $2d$ rows of $d$ slices are incomplete, $f_i=2d^2-d$. In case (5),
%% $2d$ slices are incomplete but no row is incomplete, $f_i=2d$. In case
%% (6), $d$ rows of $d$ slices are incomplete, and $d$ additional slices
%% are incomplete, $f_i$=$d^2+d-1$. And in case (7), $d$ rows of $2d$
%% slices are incomplete, $f_i$=$2d^2-1$.
In general:
\begin{equation}
  N_\mathrm{cluster\_reads}= n + \ceil*{\frac{bR}{m}}O(d^2)
\end{equation}

@@

If $m\geq\frac{Rb}{\sqrt[3]{n}^2}$, i.e., a
full row of blocks fits in memory, then a
variant adjusts the memory load to the
highest value for which only full rows are written. In this
case, $f_i \in \{1,d,2d\}$ and:
\begin{equation}
  N_\mathrm{cluster\_reads\_v1}= n + \frac{bR}{m|}O(d)
\end{equation}
where
$m|=\frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}}$. It
should be noted that $m|$ is lower than $m$, therefore
$N_\mathrm{cluster\_reads\_v1}$ might be greater than
$N_\mathrm{cluster\_reads}$ for small values of $d$.

If $m\geq\frac{Rb}{\sqrt[3]{n}}$, i.e., a full slice of blocks
fits in memory, then a second variant adjusts
the memory load to the highest value for which only full
slices are written. In this case:
\begin{equation}
  N_\mathrm{cluster\_reads\_v2}= n + \frac{bR}{m||}
\end{equation}
where $m||=\frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}$.

@@

\begin{algorithm}[h]
  \caption{Buffered merging from blocks using multiple reads
    \todo{improve notations and algo}.}
  \label{algo:multiple-reads}
  \begin{algorithmic}
  \STATE sorted\_split\_name\_list = sort blocks by increasing (x,y,z) values 
  \STATE start\_index = 0, end\_index=(m-1)
  \WHILE{end\_index \textless sizeof(reconstructed\_image)}
    \STATE initialize data\_buffer
    \FOR{split\_name in sorted\_split\_name\_list}
      \IF{indexes of split\_name is in next\_write\_range}
        \STATE read split;
        \STATE extract all indexes in next write range;
        \STATE assgin the corresponding data to the data\_buffer
      \ENDIF
    \ENDFOR
    \STATE write data\_buffer in reconstructed\_image 
    \STATE clear data\_buffer
    \STATE start\_index = end\_index + 1
    \STATE end\_index += mem
  \ENDWHILE

@@

\begin{equation}
  N_\mathrm{multiple\_reads}= n\left( \ceil*{\frac{\left((d-1)D^2+dD\right)b}{m}}\right) + 2\ceil*{\frac{bR}{m}}
\end{equation}

>>>>

\title{Sequential algorithms to split and merge ultra-high resolution 3D images}

@@

Three-dimensional images that exceed typical memory size are
increasingly found in a variety of disciplines. BigBrain, for
instance, is a 3D histological image of the human brain that
represents 1~TB of data organized in 3600 planes at full resolution
(1$\mu$m in-plane), and 76~GB at a 40$\mu$m isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. Other
examples found in medical imaging, our primary domain of interest,
include \todo{electromicroscopy,spectroscopy-francoise-peyrin}. As
such images would typically be processed on a computing cluster,
possibly using locality-aware file systems such as HDFS, software
libraries are needed to split and merge such images efficiently. We
introduce and compare a family of algorithms to address this problem
in a sequential environment made of a single storage disk. We are
aware of the need for parallel algorithms that would split and merge
images to and from an array of disks but we plan to describe such
algorithms in a subsequent paper based on the results of the present
one.

@@

We adopt the following notations:
\begin{itemize}
\item $R=D^3$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel in the reconstructed image and chunks (in B).
\item $n$: number of chunks (blocks or slices).
\item $\alpha_r$: data read rate from disk (in B/s).
\item $\alpha_w$: data write rate to disk (in B/s).
\item $\beta$: disk access time (in s).
\item $\sigma: \mathbb{R}^{+} \rightarrow \mathbb{R}^{+}$: seek time (in s), function of the seek
  distance (in B).
\item $m$: amount of available RAM (in B).
  \item $m'$: amount of used RAM (in B). $m'\leq m$.
\end{itemize}

@@

The main idea of cluster reads is to load multiple blocks in memory,
concatenate them in a buffer and write the buffer in the reconstructed
image. Seeking is reduced compared to naive block merging since
contiguous parts of the buffer will be written without seeking. In
cluster reads, a given block is accessed only once. That is, the
amount of used memory $m'$ is such that $m'=kd^3b, k \in \mathbb{N}$.

The buffer is a
slightly complex data structure, for instance an associative array or
a Python dictionary, capable of storing multiple disjoint sequences of
contiguous bytes without having to allocate memory for the bytes
between such sequences. Figure~\ref{fig:cluster-reads-buffer}
illustrates how the buffer would fill up when two blocks of one slice
and three rows per slice are loaded. Implementation details about the
buffer are given later.
\begin{figure}
\centering
\def\svgwidth{\columnwidth}
\input{figures/buffer-cluster-reads.pdf_tex}
\caption{Illustration of the buffer used in cluster reads. White
  portions are not allocated. [a,b] refers to block a, row b.}
\label{fig:cluster-reads-buffer}
\end{figure}

The number of seeks performed by cluster reads depends on how blocks
loaded in memory arrange in the reconstructed image. In the best case,
complete contiguous slices of the reconstructed image can be assembled
in memory and written without any seeking. In the worst case, some
rows in the reconstructed image are not covered by blocks loaded in
memory: O($d^2$) seeks are then required to write the memory load, one
for every partial row in every partial slice. In the intermediary
case, rows are complete but some slices can only be partially
reconstructed: O($d$) seeks are then
required. In the general case, we have:
\begin{equation}
  N_\mathrm{cluster\_reads}= n + \sum_{i=1}^{\ceil*{\frac{bR}{m}}}f_i
\end{equation}
where $f_i$ is the number of seeks done in the $i^{th}$ memory
load. Figure~\ref{fig:cluster-reads} lists the possible block
arrangements and the corresponding values of $f_i$. \todo{Val\'erie:
  Check that ordering in the Figure is consistent with Nifti's.}  From
this first analysis, we conjecture that incomplete block rows and
incomplete block slices should be avoided whenever possible. That is,
our implementation of cluster reads will consider only three cases:
\begin{itemize}
  \item Case 1: $m<d^3\sqrt[3]{n}b=\frac{D^3}{\sqrt[3]{n}^2b}$, that is, a
    row of blocks does not fit in memory. Incomplete rows of
    blocks will be merged, with complexity O($d^2$).
  \item Case 2: $\frac{D^3}{\sqrt[3]{n}^2}b\leq m<\frac{D^3}{\sqrt[3]{n}}b$,
    that is, a row of blocks fits in memory but a slice of blocks does
    not. Only complete rows of blocks will be merged, with
    complexity O(d). We have $m'=\floor*{\frac{m}{d^3\sqrt[3]{n}b}}$.
    \item Case 3: $\frac{D^3}{\sqrt[3]{n}}b \leq m < D^3b$, that is, a slice
      of blocks fits in memory. Only complete slices of blocks
      will be merged, with complexity O(1). We have
      $m'=\floor*{\frac{m}{d^3\sqrt[3]{n}^2b}}$.
\end{itemize}

Algorithm~\ref{algo:cluster-reads} shows the resulting cluster reads
method. Function \texttt{adjust\_memory} (line 4) returns $m'$
according to the three cases above. Function \texttt{sizeof} (line 6)
returns the memory used by the buffer, including only its allocated
segments. It should be noted that writing the buffer to the
reconstructed image (line 11) might involve some manipulation in the
buffer to merge its continuous segments. 
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}[1]
    \STATE list = sort blocks by increasing (x,y,z) values
    \STATE initialize buffer
    \STATE index = 0
    \STATE m'=adjust\_memory(m)    
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m'}
        \STATE block = list[index]
        \STATE read block and insert it in buffer
        \STATE index += 1
        \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

%% In case (1) on
%% this Figure, the memory load completely fills rows and slices in the
%% reconstructed image. Thus it can be written with a single seek,
%% $f_i$=1. In case (2), all the rows are complete but $d$ slices are
%% incomplete; one seek is required for each slices, $f_i=d$. In case
%% (3), $d$ rows of $d$ slices are incomplete, $f_i=d^2$. In case (4),
%% $2d$ rows of $d$ slices are incomplete, $f_i=2d^2-d$. In case (5),
%% $2d$ slices are incomplete but no row is incomplete, $f_i=2d$. In case
%% (6), $d$ rows of $d$ slices are incomplete, and $d$ additional slices
%% are incomplete, $f_i$=$d^2+d-1$. And in case (7), $d$ rows of $2d$
%% slices are incomplete, $f_i$=$2d^2-1$.

@@

The number of seeks performed by cluster reads depends on the three
cases listed above. In case 1, $f_i$ might be (a) $d^2$
 or (b) $2d^2$
depending whether the memory load
overlaps on multiple block slices, which will happen
$\alpha$
times where:
\begin{equation*}
  \alpha=\sqrt[3]{n}-1-\floor*{\frac{n-\sqrt[3]{n}^2}{LCM\left(\frac{m'}{d^3},\sqrt[3]{n}^2\right)}}
\end{equation*}
 We have:
\begin{equation}
  N_\mathrm{cluster\_reads\_v1}= n + \frac{bR}{m'}d^2+d^2\alpha
\end{equation}
Similarly, in Case 2 we have:
\begin{equation}
  N_\mathrm{cluster\_reads\_v2}= n + \frac{bR}{m'}d+d\alpha
\end{equation}
And in Case 3:
\begin{equation}
  N_\mathrm{cluster\_reads\_v3}= n + \frac{bR}{m'}
\end{equation}

@@

\begin{algorithm}[h]
  \caption{Buffered merging from blocks using multiple reads
    \todo{improve notations and algo}.}
  \label{algo:multiple-reads}
  \begin{algorithmic}
  \STATE sorted\_split\_name\_list = sort blocks by increasing (x,y,z) values 
  \STATE start\_index = 0, end\_index=(m-1)
  \WHILE{end\_index \textless sizeof(reconstructed\_image)}
    \STATE initialize data\_buffer
    \FOR{split\_name in sorted\_split\_name\_list}
      \IF{indexes of split\_name is in next\_write\_range}
        \STATE read split;
        \STATE extract all indexes in next write range;
        \STATE write the corresponding data to the data\_buffer
      \ENDIF
    \ENDFOR
    \STATE write data\_buffer in reconstructed\_image 
    \STATE clear data\_buffer
    \STATE start\_index = end\_index + 1
    \STATE end\_index += mem
  \ENDWHILE

@@

In multiple reads, a given chunk will be accessed $k$ times and we have: 
\begin{equation*}
   N_\mathrm{multiple\_reads}= nk + \ceil*{\frac{bR}{m}}
\end{equation*}
$k$ depends on the span of memory offsets covered by a chunk in the
reconstructed image, that is, $\left[(d-1)D^2+dD\right]b$.

Finally:
\begin{equation}
   N_\mathrm{multiple\_reads}= n\left( \ceil*{\frac{\left((d-1)D^2+dD\right)b}{m}}\right) + 2\ceil*{\frac{bR}{m}}
\end{equation}
